---
title: "20200210_Lin_Jiacheng_HW2"
author: "Jiacheng Lin"
date: "2020/2/10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```
## Part 1: Get Heart Disease Data Set
### Question 1  
#### *Download Heart Disease data set from UCI Machine Learning Repository*
```{bash eval = FALSE}
wget https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data
```
### Question 2
#### *Import the file into R and label the column names properly. Note you can get some necessary information from  <https://archive.ics.uci.edu/ml/datasets/Heart+Disease>. Print the first few lines of the dataframe including your header row.*  
The 14 attributes used according to the dataset website  

1. #3 (age) age in years  
2. #4 (sex) sex (1 = male; 0 = female)  
3. #9 (cp) chest pain  
* Value 1: typical angina  
* Value 2: atypical angina  
* Value 3: non-anginal pain  
* Value 4: asymptomatic  
4. #10 (trestbps) resting blood pressure (in mm Hg on admission to the hospital)  
5. #12 (chol) serum cholestoral in mg/dl  
6. #16 (fbs) (fasting blood sugar > 120 mg/dl) (1 = true; 0 = false)  
7. #19 (restecg) resting electrocardiographic results  
* -- Value 0: normal   
* -- Value 1: having ST-T wave abnormality (T wave inversions and/or ST elevation or depression of > 0.05 mV)  
* -- Value 2: showing probable or definite left ventricular hypertrophy by Estes' criteria  
8. #32 (thalach) maximum heart rate achieved  
9. #38 (exang) exercise induced angina (1 = yes; 0 = no)  
10. #40 (oldpeak) ST depression induced by exercise relative to rest  
11. #41 (slope)  the slope of the peak exercise ST segment  
* -- Value 1: upsloping  
* -- Value 2: flat  
* -- Value 3: downsloping  
12. #44 (ca) number of major vessels (0-3) colored by flourosopy  
13. #51 (thal) 3 = normal; 6 = fixed defect; 7 = reversable defect  
14. #58 (num) (the predicted attribute) num: diagnosis of heart disease (angiographic disease status)  
* -- Value 0: < 50% diameter narrowing  
* -- Value 1: > 50% diameter narrowing 
```{r}
heart_disease <- read.csv('processed.cleveland.data', header = FALSE)
colnames(heart_disease) <- c("age","sex","cp","trestbps","chol","fbs","restecg","thalach","exang","oldpeak","slope","ca","thal","num")
head(heart_disease, 5)
```
### Question 3
#### *The last column of the data contains diagnosis information ranging from 0 to 4. The “0” label indicates
that the subject is healthy, whereas the subjects with non-zero value in this field are positively
diagnosed with heart disease. Using *dplyr*, add a new column named “diagnosed” that contains binary
information about whether the subject has heart disease or not (e.g. true/false, 0/1, etc). Print the first
few rows of your dataframe to confirm you have successfully created the new column.*  
```{r}
heart_disease <- heart_disease %>% mutate(diagnosed = ifelse(num > 0, TRUE, FALSE))
head(heart_disease)
```
## Part 2: Explore and Visualize Heart Disease Data Set
### Question 1
#### *Summarize the data. Your summary should include the following information below.*
##### *i. number of counts by diagnosis status (use “diagnosed” column info)*
```{r}
positive = summarise(heart_disease, positive = sum(diagnosed == TRUE))
negative <- summarise(heart_disease, negative = sum(diagnosed == FALSE))
cat("Positive diagnosed patients number is", positive[1,1], "\n")
cat("Negative diagnosed patients number is", negative[1,1])
```
Positive diagnosed patients number is 139, and Negative diagnosed patients number is 164.
##### *ii. number of counts by sex*
```{r}
male = summarise(heart_disease, male = sum(diagnosed == 1))
female <- summarise(heart_disease, female = sum(diagnosed == 0))
cat("Male patients number is", male[1,1], "\n")
cat("Female patients number is", female[1,1])
```
Male patients number is 139 and female patients number is 164.
##### *iii. average of cholesterol level by diagnosis status (use “diagnosed” column info)*
```{r}
positive_chol = summarise(heart_disease, chol = mean(chol[diagnosed == TRUE]))
negative_chol = summarise(heart_disease, chol = mean(chol[diagnosed == FALSE]))
cat("Positive diagnosed patients average cholesterol is", positive_chol[1,1], "\n")
cat("Negative diagnosed patients average cholesterol is", negative_chol[1,1])
```
Positive diagnosed patients average cholesterol is 251.4748 mg/dl 
and Negative diagnosed patients average cholesterol is 242.6402 mg/dl
##### *iv. min and max of the subjects’ age by diagnosis status (use “diagnosed” column info)*
```{r}
age_pos = summarise(heart_disease, old = max(age[diagnosed == TRUE]), young = min(age[diagnosed == TRUE]))
age_neg = summarise(heart_disease, old = max(age[diagnosed == FALSE]), young = min(age[diagnosed == FALSE]))
cat("The oldest positive diagnosed patient is", age_pos[1,1], 
    "and the youngest is", age_pos[1,2], "\n")
cat("The oldest negative diagnosed patient is", age_neg[1,1], 
    "and the youngest is", age_neg[1,2])
```
The oldest positive diagnosed patient is 77 and the youngest is 35.
The oldest negative diagnosed patient is 76 and the youngest is 29.
